### 2022-10-09 AISTATS2023/paper.tex
small modifications

\subsection{Heterogeneity Of Instance-Level Effects.}
\label{sec:quant-heter-effects}

@@

Heterogeneity It comes naturally to also ask \textit{to
  what extent }. Quantifying the uncertainty of the global explanation
has attracted a lot of interest. ICE and
d-ICE\citep{goldstein2015peeking} provide a set of curves that are
plot on top-of the PDP. Both methods produce one curve for each
instance of the training set;
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\) for d-ICE. The user then visually observes if the curves are
homogeneous and to what extent they deviate from the PDP. Some methods
try to automate the aforementioned visual exploration, by grouping
(d-)ICE plots into clusters~\citep{molnar2020model,
  herbinger2022repid, britton2019vine}. Some other approaches, like
H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
index\citep{greenwell2018simple} or SHAP interaction
values\citep{lundberg2018consistent}, quantify the level of
interaction between the input features, with an interaction value. A
strong interaction index is an indicator for the existence of
heterogeneous effects.

The aforementioned approaches are under two limitations; They either
do not quantify the uncertainty of the FE directly or they are based
on PDPs, and, therefore, they are subject to the failure modes of PDPs
in cases of correlated features\citep{baniecki2021fooling}, as we will
show in-depth in Section~\ref{sec:simulation-examples-1}. To the best
of our knowledge, no work exist so far that quantifies the
heterogeneous effects based on the formulation of ALE.

\subsection{ALE Approximation.}
\label{sec:ale-approximation}

In ML scenarios, ALE is estimated from the training set
examples. Therefore, \citep{apley2020visualizing} proposed dividing
the feature domain in \(K\) equal-sized bins and estimating the local effects in
each bin by evaluating the black box-function at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
(DALE) that computes the local effects on the training instances using
auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

Both approximations ask from the user to blindly decide the number of
bins, denoted with \(K\), for splitting the axis into equally-sized
bins. This approach introduces significant limitations. Specifically, setting \(K\) to a small
value may hide fine-grain effects due to large bins while setting \(K\)
to a high value leads to poor bin-effect estimations from limited
samples. In general, the user may face contradictory explanations for
different \(K\) without a clue for which one to trust.

@@

UALE extends ALE for quantifying the level of heterogeneous effects
(uncertainty) and automates the bin-splitting. At
Section~\ref{sec:UALE-definition-1} we define UALE, at
Section~\ref{sec:UALE-definition-2} we redefine it using an
interval-based formulation, at
Section~\ref{sec:UALE-approximation} we provide an
approximation and at Section~\ref{sec:bin-spliting} we define and
solve the problem of optimal bin-splitting.

@@

We extend ALE definition of Eq.~(\ref{eq:ALE}) with a component for
quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations
\(\mathcal{H}(z) = \sigma(z)\), where:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(z) = \mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] 
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
UALE formulates the effect at a specific point \(x_s\) with a compact
tuple that consists of the average effect and the uncertainty
\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve with
a confidence region, i.e.
\(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

@@

In real scenarios, all estimations are based on the instances of the
training set. Therefore, we cannot estimate \(\mu(z), \sigma(z)\) at
the granularity of a point, because the probability of observing a
sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). We define the regional-effect \(\mu(z_1, z_2)\) and the
regional-uncertainty \(\mathcal{H}(z_1, z_2) = \sigma(z_1, z_2)\), as the summary statistics
that characterize the effect and the uncertainty inside the interval
\([z_1, z_2)\):

>>>>

\subsection{Heterogeneity Of Instance-Level Effects}
\label{sec:quant-heter-effects}

@@

ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
are plot on top-of PDP. Both methods produce one curve for each
instance of the training set;
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\) for d-ICE. The user then visually observes if the curves are
homogeneous and to what extent they deviate from the PDP. Some methods
try to automate the aforementioned visual exploration, by grouping
(d-)ICE plots into clusters~\citep{molnar2020model,
  herbinger2022repid, britton2019vine}. Other approaches, like
H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
index\citep{greenwell2018simple} or SHAP interaction
values\citep{lundberg2018consistent} inform indirectly about
heterogeneous effects, quantifying interactions between the input
features. A strong interaction index is an indicator for the existence
of heterogeneous effects.

The aforementioned approaches are under two limitations. They either
do not quantify the uncertainty of the FE directly or they are based
on PDPs, and, therefore, they are subject to the failure modes of PDPs
in cases of correlated features\citep{baniecki2021fooling}, as we will
show in-depth in Section~\ref{sec:simulation-examples-1}. To the best
of our knowledge, no work exist so far that quantifies the
heterogeneous effects based on the formulation of ALE.

\subsection{ALE Approximation}
\label{sec:ale-approximation}

In ML scenarios, ALE is estimated from the instances of the training
set.~\citep{apley2020visualizing} proposed dividing the feature domain
in \(K\) equisized bins and estimating the local effects in each bin
by evaluating the black box-function at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instances of the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
(DALE) that uses the partial derivatives:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution when
bins become wide and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

Both approximations ask from the user to blindly decide the bin
density for splitting the feature axis into \(K\) equisized bins, an
approach that is error prone. Setting \(K\) to a small value may hide
fine-grain effects due to large bins while setting \(K\) to a high
value leads to poor bin-effect estimations due to few samples per
bin. In general, the user may receive contradictory explanations for
different \(K\) without a clue for which one to trust.

@@

UALE extends ALE for quantifying the level of heterogeneous effects
(uncertainty) and automates the bin-splitting. At
Section~\ref{sec:UALE-definition-1} we define UALE, at
Section~\ref{sec:UALE-definition-2} we reformulate it using an
interval-based formulation, at Section~\ref{sec:UALE-approximation} we
provide UALE approximation and at Section~\ref{sec:bin-spliting} we
define and solve the problem of optimal bin-splitting.

@@

UALE extends ALE definition of Eq.~(\ref{eq:ALE}) with a component for
quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it with the standard deviation of the local explanations
\(\mathcal{H}(z) = \sigma(z)\), where:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(z) = \mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] 
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
UALE formulates the effect at a specific point \(x_s\) with a tuple
that consists of the average effect and the uncertainty
\((\mu(z), \sigma(z))\) and visualizes it with a continuous curve and
a confidence region
\(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

@@

In real ML scenarios, all estimations are based on the instances of
the training set. Therefore, we cannot estimate \(\mu(z), \sigma(z)\)
at the granularity of a point, because the probability of observing a
sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). We define the regional-effect \(\mu(z_1, z_2)\) and the
regional-uncertainty \(\mathcal{H}(z_1, z_2) = \sigma(z_1, z_2)\), as
the summary statistics that characterize the effect and the
uncertainty inside the interval \([z_1, z_2)\):
